---
title: "Session 2: LLMs"
format:
  html:
    toc: true
    toc-depth: 2
execute:
  warning: false
  message: false
---

## Overview

In this session we introduce large language models (LLMs) and explore how they can be used in ecological and environmental applications. We begin by accessing commercial and open-source models, then move on to repeated prompting, model comparison, and a press release example based on a scientific paper.

[Download the R script](files/session-2-llms.R)

## Packages

```{r}
#| eval: true
#| echo: false
require(ellmer)
require(ollamar)
library(pdftools)
```

## 1. Accessing the OpenAI API

We are going to use the `ellmer` package to access LLMs. First we will use a commercial model provided by OpenAI and ask for an opinion about New Zealand's sustainability.

To access an OpenAI model, you will need an API key. A safe way to do this is to set it outside the script as an environment variable.

Example:

```{r}
#| eval: false
Sys.setenv(OPENAI_API_KEY = "insert your key here")
```

We can then open a chat using the `chat_openai()` function and send a question to the model.

```{r}
#| eval: true
#| echo: true
newchat <- chat_openai(
  model = "gpt-4o",
  system_prompt = "You are an expert on global sustainability."
)

res <- newchat$chat(
  "How sustainable is New Zealand's economy?",
  echo = "none"
)

print(res)
```

## 2. Accessing an open-source local model

We can also try an open-source alternative using the `ollamar` package. This requires a little more setup because we first need to install a local model.

```{r}
#| eval: true
#| echo: true
list_models()
```

Let's install a really tiny one so that it can run quickly on a normal laptop.

```{r}
#| eval: true
#| echo: true
pull("tinyllama:1.1b")
```

Once a model has been installed, we can access it in a similar way to the OpenAI model.

```{r}
#| eval: true
#| echo: true
newchat <- chat_ollama(
  model = "tinyllama:1.1b",
  system_prompt = "You are an expert on global sustainability."
)

res <- newchat$chat(
  "How sustainable is New Zealand's economy?",
  echo = "none"
)

print(res)
```

## 3. Asking a series of questions

In environmental social science we may be interested in asking land use questions to virtual "agents". This is increasingly common in simulation studies and agent-based modelling.

Here we consider a situation where a farmer might change part of their farm from sheep farming to native forestry. Whether or not the farmer plants trees may depend on the size of the financial incentive available.

```{r}
#| eval: true
#| echo: false
dollars <- seq(0,500, by = 40)

agentq <- data.frame(
  dollars = dollars,
  chatgpt = NA,
  open = NA
)

systemtext <- "You are a New Zealand sheep farmer making a land use decision."

qulist <- paste0(
  "You are a sheep farmer in New Zealand, with 150 hectares of land. ",
  "You make an average profit of $438 per hectare from farming per year. ",
  "You are making a decision about whether to plant native trees on your land instead of farming sheep next year. ",
  "The cost of planting native trees is $100 per hectare and you would lose the annual profit from sheep farming. ",
  "The government has offered to pay you $",
  dollars,
  " per hectare year if you plant the native trees. Please make this decision while thinking about your livelihood. ",
  "Please choose how many hectares of land to convert to native trees by indicating the number of hectares. ",
  "Do not provide any other text or response, only a number between 0 and 150 indicating the number of hectares."
)
```

Next we send these prompts to the OpenAI model.

```{r}
#| eval: true
#| echo: false
for (j in 1:length(qulist)) {

  farmerchat <- chat_openai(
    model = "gpt-4o",
    system_prompt = systemtext
  )

  res <- farmerchat$chat(qulist[j], echo = "none")
  agentq[j,2] <- as.numeric(res)
}
```

If the responses are collected successfully, we can plot the results.

```{r}
#| eval: true
#| echo: true
plot(chatgpt ~ dollars,
     pch = 16,
     col = "red",
     data = agentq,
     xlab = "Incentive ($)",
     ylab = "Forested area (ha)")

m1 <- glm(cbind(chatgpt,150-chatgpt) ~ dollars,
          data = agentq,
          family = "binomial")

lines(1:500,
      predict(m1,
              newdata = data.frame(dollars = 1:500),
              type = "response")*150,
      lty = 3)
```

So this is cool! The simulated farmer is making economically-driven decisions, displaying some financial logic. At a low incentive level, they choose to keep their farms in productive agriculture. However, if they are provided with a large enough incentive, they will switch their behaviour and start planting trees. You can read more about this topic in our recent [paper](https://iopscience.iop.org/article/10.1088/2515-7620/ae4cfb).

## 4. Repeating the survey with an open-source model

We can repeat the same loop using a local model.

```{r}
#| eval: true
#| echo: false
for (j in 1:length(qulist)) {

  farmerchat <- chat_ollama(
    model = "tinyllama:1.1b",
    system_prompt = systemtext
  )

  res <- farmerchat$chat(qulist[j], echo = "none")
  agentq[j,3] <- as.numeric(res)
}
```

We can inspect the responses.

```{r}
#| eval: true
#| echo: true
agentq$open
```

Oh dear, the model has not been able to perform the task properly, likely because it is not complex enough. Let's see what it had to say;

```{r}
#| eval: true
#| echo: true
agentq$open
res
```

It is actually trying to provide some logic, but unfortunately has not been able to appreciate the request or particularly the importance of providing a simple numerical response.

## 5. Press release example

LLMs can also be used for science communication. In this example we extract the text from a PDF and ask the model to prepare a press release. You can use any pdf file that you like for this test. I used a recent [paper](https://besjournals.onlinelibrary.wiley.com/doi/abs/10.1002/pan3.10622) about the potential of generative AI for climate change adaptation.

```{r}
#| eval: true
#| echo: false
pdf_path <- "C:/Users/RichardsD/OneDrive - MWLR/Documents/GithubRepo/aicourse/Richards et al 2024.pdf"
papertext <- pdftools::pdf_text(pdf_path)
```

```{r}
#| eval: false
#| echo: true
pdf_path <- "Data/Richards et al 2024.pdf"
papertext <- pdftools::pdf_text(pdf_path)
```

We then ask the model to write a press release.

```{r}
#| eval: true
#| echo: true
newchat <- chat_openai(
  model = "gpt-4o",
  system_prompt = "You are an expert in writing press releases for scientific papers."
)

res <- newchat$chat(
  paste0(
    "Please read the following paper and prepare a press release;\n\n",
    papertext,
    "\nThe press release should be less than 500 words."
  ),
  echo = "none"
)

print(res)
```

We can also request a social media post.

```{r}
#| eval: true
#| echo: true
res2 <- newchat$chat(
  "Thanks, please could you now write a suitable social media post e.g. for LinkedIn?",
  echo = "none"
)

print(res2)
```

In this session we used both commercial and open-source LLMs, explored repeated prompting in a land-use decision example, and used an LLM to help communicate a scientific paper. These examples show both the usefulness of LLMs and some of their practical limitations.
